---
title: "Example code"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=TRUE, eval = F}
knitr::opts_chunk$set(echo = TRUE)
library(INLA)
library(lme4)
```

This document contains example code used to conduct the analysis in "A Bayesian hierarchical model for predicting rates of oxygen consumption in mechanically ventilated intensive care patients".

# Data

The data contains breath-by-breath measurements from 35 patients (pid) across over 100 rehabilitation sessions (vid). We are primarily interested in predicting $\dot VO_2$ (`VO2_kg`).

# Bayesian hierarchical models

The fully specified Bayesian hierarchical model defined in the paper can be run as follows

```{r, eval = F}
######## Fully specified model
# Prior for the O-U process
a1 <- 4e+6
b1 <- 1e+3
prec.prior.ou <- list(prec = list(prior = "loggamma", 
                               param = c(a1, b1)
                               )
                   )
# Priors for fixed effects
prior.fixed <- list(prec = 0.1, prec.intercept = 0.1)
# Covariates placed on log scale and centered beforehand
formula <- log(VO2_kg) ~ clogVt + clogPETCO2 + clogRR +
  calc_sofa + sex + clogage + GPPAQ + clogBMI + clogage*clogBMI +
  clogVt*clogRR + clogVt*clogPETCO2 +
  f(numvid, model = "iid") + # Random intercept
  f(numpid, clogVt, model = "iid") + # Random slope
  f(Time.1, # Temporal effect
    model = "ou", # OU process smoothing term
    group = numvid, # Group by session 
    control.group = list(model = "iid"),
    hyper = list(prec = prec.prior.ou)
    )  # Dependence between tests

mod_full  <-  inla(formula,
               data = dat,
               control.fixed = prior.fixed,
               control.predictor = list(compute = TRUE),
               control.compute = list(config = TRUE,
                                      dic = T)
             )
### Get model output
summary(mod1)
print(mod1$summary.hyperpar)

### For the O-U only model 
prec.prior1 <- list(prec = list(initial = 15,
                                fixed = TRUE
                               )
                   )

mod_OU  <-  inla(formula,
               data = dat,
               control.family = list(hyper = prec.prior1),
               control.fixed = prior.fixed,
               control.predictor = list(compute = TRUE),
               control.compute = list(config = TRUE,
                                      dic = T)
               )
               
#### For the iid error model
formula <- log(VO2_kg) ~ clogVt + clogPETCO2 + clogRR +
  calc_sofa + sex + clogage + GPPAQ + clogBMI + clogage*clogBMI +
  clogVt*clogRR + clogVt*clogPETCO2 +
  f(numvid, model = "iid") + # Random intercept
  f(numpid, clogVt, model = "iid")

mod_full  <-  inla(formula,
               data = dat,
               control.fixed = prior.fixed,
               control.predictor = list(compute = TRUE),
               control.compute = list(config = TRUE,
                                      dic = T)
             )
```



# Posterior distribution samples

The following code can be used to generate samples from the posterior distribution.

```{r, eval = F}
n <- # Number of samples
mod <- # INLA model output as previously specified
Sample <- inla.posterior.sample(n, mod,
                               selection = list(`(Intercept)` = 1,
                                                `clogVt` = 1,
                                                `clogPETCO2` = 1,
                                                `clogRR` = 1,
                                                `calc_sofa` = 1,
                                                `sex1` = 1,
                                                `clogage` = 1,
                                                `GPPAQ2` = 1,
                                                `GPPAQ3` = 1,
                                                `GPPAQ4` = 1,
                                                `clogBMI` = 1,
                                                `clogVt:clogRR` = 1,
                                                `clogVt:clogPETCO2` = 1,
                                                `clogage:clogBMI` = 1
                               ))
# Generate a predictive distribution sample
cov <- # vector of covariate values
coef <- Samples[[1]]$latent # Fixed effect samples
var_ <- Samples[[1]]$hyperpar[[1]]^(-1) + Samples[[1]]$hyperpar[[2]]^(-1) + 
  # Sample specifically for random slope value
  clogVt*Samples[[3]]$hyperpar[[1]]^(-1) + 
  Samples[[4]]$hyperpar[[1]]^(-1)
# Generate sample
pred_sample <- coef*cov + rnorm(1, mean = 0, sd = sqrt(var_))
```

# OUES model

The OUES model is fit using the \texttt{lme4} R package.

```{r, eval = FALSE}
m_OUES <- lmer(VO2_kg ~ (1|pid) + log10(VE) + (0+log10(VE)|pid), data = dat)
```
