---
title: "Example code"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=TRUE, eval = F}
knitr::opts_chunk$set(echo = TRUE)
library(INLA)
library(lme4)
```

This document contains example code used to conduct the analysis in "A Bayesian hierarchical model for predicting rates of oxygen consumption in mechanically ventilated intensive care patients".

# Data

The data contains breath-by-breath measurements from 35 patients (pid) across over 100 rehabilitation sessions (vid). We are primarily interested in predicting $\dot VO_2$ (`VO2_kg`).

A simulated dataset code be generate using the following code:

```{r, eval = F}
n <- 100
s <- rep(NA,n)
clogVT <- rep(NA,n)
clogPET <- rep(NA,n)
clogRR <- rep(NA,n)
logVO2_kg <- rep(NA,n)
t <- rep(NA,n)
s <- rep(NA,n)
pid <- rnorm(1, mean = 0, sd = sqrt(1/48.39))
vid <- rnorm(1, mean = 0, sd = sqrt(1/31.07))
# User input values
sex <- 0 
clogage <- 0
clogBMI <- 0
sofa <- 10 # Sepsis organ failure assessment score
GPPAQ <- 1 # 1-4

for(i in 1:n){
  if(i == 1){
    t[i] <- 0
    # OU process
    s[i] <- rnorm(1, 0 , sd = sqrt(1/41))
  }else{
    # Time between breathes
    t[i] <- t[i-1] + rexp(1) 
    s[i] <- rnorm(1, mean = s[i-1]*exp(-(t[i] - t[i-1])*0.04), sd = sqrt(1/(41*(1-exp(-2*(t[i] - t[i-1])*0.04))^(-1))))
  }
  # Physiological values (centered and on log scale) - doesn't reflect correlation structure we see in the data but can be used to reproduce results
  clogVT[i] <- rnorm(1, mean = 0, sd = 2)
  clogPET[i] <- rnorm(1, mean = 0, sd = 0.5)
  clogRR[i] <- rnorm(1, mean =, sd = 0.2)
  logVO2_kg[i] <- 1.33 + pid + (1.52 + vid)*clogVT[i] + 1.91*clogPET[i] + 1.09*clogRR[i] -0.11*(clogVT[i]*clogPET[i]) + 0.3*(clogVT[i]*clogRR[i]) -
    0.01*sofa - 0.01*(GPPAQ == 2) + 0.01*(GPPAQ == 3) + 0.29*(GPPAQ == 4) -0.07*sex + 0.34*clogage - 1.12*clogBMI -2.06*clogage*clogBMI + s[i] +rnorm(1,0,sd = sqrt(1/256.72))
}

dat <- data.frame(t, logVO2_kg, clogVT, clogPET,clogRR, sex, clogage, clogBMI, sofa, GPPAQ)


```


# Bayesian hierarchical models

The fully specified Bayesian hierarchical model defined in the paper can be run as follows

```{r, eval = F}
######## Fully specified model
# Prior for the O-U process
a1 <- 4e+6
b1 <- 1e+3
prec.prior.ou <- list(prec = list(prior = "loggamma", 
                               param = c(a1, b1)
                               )
                   )
# Priors for fixed effects
prior.fixed <- list(prec = 0.1, prec.intercept = 0.1)
# Covariates placed on log scale and centered beforehand
formula <- log(VO2_kg) ~ clogVt + clogPETCO2 + clogRR +
  calc_sofa + sex + clogage + GPPAQ + clogBMI + clogage*clogBMI +
  clogVt*clogRR + clogVt*clogPETCO2 +
  f(numvid, model = "iid") + # Random intercept
  f(numpid, clogVt, model = "iid") + # Random slope
  f(Time.1, # Temporal effect
    model = "ou", # OU process smoothing term
    group = numvid, # Group by session 
    control.group = list(model = "iid"),
    hyper = list(prec = prec.prior.ou)
    )  # Dependence between tests

mod_full  <-  inla(formula,
               data = dat,
               control.fixed = prior.fixed,
               control.predictor = list(compute = TRUE),
               control.compute = list(config = TRUE,
                                      dic = T)
             )
### Get model output
summary(mod1)
print(mod1$summary.hyperpar)

### For the O-U only model 
prec.prior1 <- list(prec = list(initial = 15,
                                fixed = TRUE
                               )
                   )

mod_OU  <-  inla(formula,
               data = dat,
               control.family = list(hyper = prec.prior1),
               control.fixed = prior.fixed,
               control.predictor = list(compute = TRUE),
               control.compute = list(config = TRUE,
                                      dic = T)
               )
               
#### For the iid error model
formula <- log(VO2_kg) ~ clogVt + clogPETCO2 + clogRR +
  calc_sofa + sex + clogage + GPPAQ + clogBMI + clogage*clogBMI +
  clogVt*clogRR + clogVt*clogPETCO2 +
  f(numvid, model = "iid") + # Random intercept
  f(numpid, clogVt, model = "iid")

mod_full  <-  inla(formula,
               data = dat,
               control.fixed = prior.fixed,
               control.predictor = list(compute = TRUE),
               control.compute = list(config = TRUE,
                                      dic = T)
             )
```



# Posterior distribution samples

The following code can be used to generate samples from the posterior distribution.

```{r, eval = F}
n <- # Number of samples
mod <- # INLA model output as previously specified
Sample <- inla.posterior.sample(n, mod,
                               selection = list(`(Intercept)` = 1,
                                                `clogVt` = 1,
                                                `clogPETCO2` = 1,
                                                `clogRR` = 1,
                                                `calc_sofa` = 1,
                                                `sex1` = 1,
                                                `clogage` = 1,
                                                `GPPAQ2` = 1,
                                                `GPPAQ3` = 1,
                                                `GPPAQ4` = 1,
                                                `clogBMI` = 1,
                                                `clogVt:clogRR` = 1,
                                                `clogVt:clogPETCO2` = 1,
                                                `clogage:clogBMI` = 1
                               ))
# Generate a predictive distribution sample
cov <- # vector of covariate values
coef <- Samples[[1]]$latent # Fixed effect samples
var_ <- Samples[[1]]$hyperpar[[1]]^(-1) + Samples[[1]]$hyperpar[[2]]^(-1) + 
  # Sample specifically for random slope value
  clogVt*Samples[[3]]$hyperpar[[1]]^(-1) + 
  Samples[[4]]$hyperpar[[1]]^(-1)
# Generate sample
pred_sample <- coef*cov + rnorm(1, mean = 0, sd = sqrt(var_))
```

# OUES model

The OUES model is fit using the \texttt{lme4} R package.

```{r, eval = FALSE}
m_OUES <- lmer(VO2_kg ~ (1|pid) + log10(VE) + (0+log10(VE)|pid), data = dat)
```
